---
title: "Getting Started with chef"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with chef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `{chef}` package provides a framework for generating statistical evidence from clinical trial data. It helps you define endpoints and automatically generate statistical analyses using the `{targets}` pipeline system, enabling reproducible and automated evidence generation workflows.

`{chef}` is particularly useful for healthcare technology assessments (HTA) and regulatory submissions where reproducibility and auditability are critical.

## Getting Started

### Installation

You can install `{chef}` from CRAN:

```{r install}
install.packages("chef")
```

Or the development version from GitHub:

```{r install-dev}
# install.packages("devtools")
devtools::install_github("hta-pharma/chef")
```

### Basic Workflow

The typical workflow with `{chef}` involves three main steps:

1. **Setup your project** with `use_chef()`
2. **Define your endpoints** in the configuration files
3. **Run the pipeline** with `run_pipeline()`

### Example: Setting Up a Chef Project

```{r setup-project}
library(chef)

# Create a new chef project in your working directory
# This creates the necessary template files and directory structure
use_chef(
  path = "my_analysis",
  overwrite = FALSE
)
```

This creates:
- `R/` directory for endpoint and function definitions
- `_targets.R` for the targets pipeline configuration
- Template functions for data preparation and statistical analysis
- Example endpoint definitions

### Defining Endpoints

Endpoints are defined using the `define_expanded_ep()` function. Each endpoint specifies:
- Endpoint name and label
- Population criteria (e.g., safety population)
- Event filters and grouping
- Stratification variables
- Statistical functions to apply

Example endpoint definition:

```{r define-endpoints}
library(data.table)

# Define a simple endpoint
my_endpoints <- define_expanded_ep(
  endpoint_spec_id = 1,
  endpoint_label = "Primary Efficacy",
  pop_var = "SAFFL",
  pop_value = "Y",
  stratify_by = list(c("SEX")),
  treatment_var = "TRT01A"
)
```

### Running the Pipeline

Once your endpoints and statistical functions are defined, run the pipeline:

```{r run-pipeline}
# Execute the full pipeline
run_pipeline(
  adam_list = list(adae = adae_data, adcm = adcm_data),
  endpoints = my_endpoints
)
```

This automatically:
1. Prepares your clinical data
2. Applies endpoint criteria
3. Generates stratified analyses
4. Creates statistical outputs

## Key Functions

- `use_chef()` - Initialize a new chef project
- `define_expanded_ep()` - Define endpoint specifications
- `expand_over_endpoints()` - Expand endpoints across stratification levels
- `add_event_index()` - Map endpoint criteria to data rows
- `apply_criterion_endpoint()` - Apply inclusion/exclusion criteria
- `apply_stats()` - Apply statistical functions
- `run_pipeline()` - Execute the full analysis pipeline
- `stage_pipeline()` - Prepare pipeline for execution

## Documentation and Examples

For more detailed information on specific functions, see the function documentation:

```{r help}
?use_chef
?run_pipeline
?define_expanded_ep
```

## Further Resources

- GitHub: https://github.com/hta-pharma/chef
- Documentation: https://hta-pharma.github.io/chef/
- {targets} documentation: https://books.ropensci.org/targets/

## Contact and Support

For questions, bug reports, or feature requests, please visit:
https://github.com/hta-pharma/chef/issues
